---
title: "synthetic control: elon's tweet tanked tesla's stock"
subtitle: |
  a terse analysis of tesla stock prices and how one of elon's tweets moves them
date: "2020-05-01"
execute:
  echo: false
categories: [data analysis]
---

### Summary

how to use `vsp`. at the end, visualization and results takeaway for non-technical folks

###

```{r}
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(here)

path <- here("post", "xxxx-xx-xx_a-cocktail-data-analysis", "boston_cocktails.csv")

# boston_cocktails <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-26/boston_cocktails.csv")
# 
# boston_cocktails |>
#   write_csv(path)

boston_cocktails <- read_csv(path)

boston_cocktails %>%
  count(ingredient, sort = TRUE)


```

```{r}
# old mr boston ingredients really fucking shit up, my guess is that this a collection of cocktails designed to show off their products
# furthers my impression the cocktails themselves are likely to be ass
# as always, touch data/grass
# data cleaning challenge: some cocktails call for specific gins, some for a style of gin, some for "gin"
# data cleaning note: discarding infrequent ingredients is a morally incorrect thing to do with cocktails
# data cleaning note: cluster based on what? flavor? spec? presence/absence vs ratio vs volume

cocktails_parsed <- boston_cocktails %>%
  mutate(
    ingredient = str_to_lower(ingredient),
    ingredient = str_replace_all(ingredient, "-", " "),
    ingredient = str_remove(ingredient, " liqueur$"),
    ingredient = str_remove(ingredient, "old mr. boston "),
    ingredient = str_remove(ingredient, "mr. boston "),
    ingredient = str_remove(ingredient, " flavored"),
    ingredient = str_remove(ingredient, " (if desired)$"),
    ingredient = case_when(
      str_detect(ingredient, "bitters") ~ "bitters",
      str_detect(ingredient, "lemon") ~ "lemon juice",
      str_detect(ingredient, "lime") ~ "lime juice",
      str_detect(ingredient, "grapefruit") ~ "grapefruit juice",
      str_detect(ingredient, "orange") ~ "orange juice",
      TRUE ~ ingredient
    ),
    measure = case_when(
      str_detect(ingredient, "bitters") ~ str_replace(measure, "oz$", "dash"),
      TRUE ~ measure
    ),
    measure = str_replace(measure, " ?1/2", ".5"),
    measure = str_replace(measure, " ?3/4", ".75"),
    measure = str_replace(measure, " ?1/4", ".25"),
    measure_number = parse_number(measure),
    measure_number = if_else(str_detect(measure, "dash$"),
      measure_number / 50,
      measure_number
    )
  ) %>%
  # filter here probably fucks things up!
  add_count(ingredient) %>%
  # filter(n > 15) %>%
  # select(-n) %>%
  distinct(row_id, ingredient, .keep_all = TRUE) %>%
  na.omit()

cocktails_parsed |> 
  distinct(ingredient) |> 
  arrange(ingredient)

cocktails_parsed
```

```{r}
cocktails_df <- cocktails_parsed %>%
  select(-ingredient_number, -row_id, -measure) %>%
  pivot_wider(names_from = ingredient, values_from = measure_number, values_fill = 0) %>%
  janitor::clean_names() %>%
  na.omit()

cocktails_df
```

> The biggest difference in PC1 is powdered sugar vs. simple syrup; recipes are not likely to have both, which makes sense! Let’s zoom in on the first four components, and understand which cocktail ingredients contribute in the positive and negative directions.

```{r}
library(tidygraph)
library(tidytext)
library(vsp)
library(furrr) # hack for vsp bug: FIX

cocktails_parsed

A <- cast_sparse(cocktails_parsed, row = name, column = ingredient, value = measure_number)
A <- sign(A)

fa <- vsp(A, rank = 20, degree_normalize = FALSE, center = TRUE)
fa

s <- localization_statistics(A, 20)
plot_cumulative_curves(s) +
  geom_vline(xintercept = s$avg_row_sum) +
  geom_vline(xintercept = s$avg_col_sum)


# TODO: clarify the idea of "component eigen-swapping" and point out that jitters in the localization plot probably indicate eigen-swapping
```

```{r}
plot_ipr_pairs(fa)
plot_mixing_matrix(fa)
plot_varimax_y_pairs(fa, 1:10)
plot_varimax_z_pairs(fa, 1:10)
```

```{r}
screeplot(fa)
```

```{r}
get_y_hubs(fa, hubs_per_factor = 3)
```

```{r}
# Z is sparse: drinks typically only load on a few ingredients at a time
image(fa$Z)
```

```{r}
z_names <- fa |> 
  get_z_hubs(1) |> 
  pull(id)

y_names <- fa |> 
  get_y_hubs(1) |> 
  pull(id)

fa_named <- fa |> 
  set_y_factor_names(y_names) |> 
  set_z_factor_names(z_names)

plot_mixing_matrix(fa_named) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = .5)
  )
```

```{r}
fa |> 
  get_y_hubs(3) |> 
  mutate(id = reorder_within(id, loading, factor)) %>%
  ggplot(aes(loading, id, fill = loading > 0)) +
  geom_col() +
  facet_wrap(~factor, scales = "free_y") +
  scale_y_reordered() +
  labs(
    x = "Absolute value of contribution",
    y = NULL, fill = "Positive?"
  )
```

> So PC1 is about powdered sugar + egg + gin drinks vs. simple syrup + lime + tequila drinks. This is the component that explains the most variation in drinks. PC2 is mostly about vermouth, both sweet and dry.

```{r}
fa |> 
  get_z_hubs(5) |> 
  mutate(id = reorder_within(id, loading, factor)) %>%
  ggplot(aes(loading, id, fill = loading > 0)) +
  geom_col() +
  facet_wrap(~factor, scales = "free_y") +
  scale_y_reordered() +
  labs(
    x = "Absolute value of contribution",
    y = NULL, fill = "Positive?"
  ) +
  theme_classic(9) +
  theme(
    legend.position = "none"
  )
```
how many ingredients in each cocktail on average? 

```{r}

## odd discovery: applejack punch doesn't have applejack in it?

fa |> 
  get_z_hubs(5) |> 
  left_join(cocktails_parsed, by = join_by(id == name)) |> 
  select(factor, id, ingredient, measure)
```


```{r}

ig <- igraph::graph_from_incidence_matrix(A, weighted = TRUE)

```

### Normalization / scaling

```{r}
library(Matrix)

L <- sign(A) # rowScale(A, 1 / rowSums(A))
faL <- vsp(L, rank = 20, degree_normalize = FALSE)

plot_ipr_pairs(faL)
plot_mixing_matrix(faL)

faL |> 
  get_y_hubs(3) |> 
  View()


faL |> 
  get_z_hubs(3) |> 
  View()
```



https://juliasilge.com/blog/cocktail-recipes-umap/